---
title: "p8105_hw2_cd3591"
author: "Anny"
date: "2025-09-29"
output: github_document
---

```{r}
# setup library
library(tidyverse)
library(readxl)
```

# Problem 1
```{r}
# read the pols-month data and clean it
pols_data = read_csv("pols-month.csv") %>%  
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%  # separate mon into year, month and day by "-"
  mutate(year = as.integer(year),  # change year (character) variable into integer
         month = as.integer(month),  # change month (character) variable into integer
         day = as.integer(day), # change day (character) variable into integer
         month = month.name[month],  # replace month number with month name 
         president = if_else(prez_gop == 1, "gop", "dem") # create president variable taking values gop and dem
         ) %>%
  select(-prez_dem, -prez_gop, -day)  # remove prez_dem and prez_gop and the day variable

```

```{r}
# read the snp data and clean it
snp_data = read_csv("snp.csv") %>%
  janitor::clean_names() %>%
  mutate(date  = mdy(date),  # make the date into standard version
         date = if_else(date > Sys.Date(), date %m-% years(100), date)) %>%   # e.g. let 2068 into 1968
  separate(date, into = c("year", "month", "day"), sep = "-") %>%  # separate date into month, day and year by "-"
  mutate(year = as.integer(year),  # change year (character) variable into integer
         month = as.integer(month),  # change month (character) variable into integer
         day = as.integer(day), # change day (character) variable into integer
         month = month.name[month]  # replace month number with month name
         ) %>%
  select(-day) %>%  # remove the day variable
  relocate (year, month)  # make sure year and month are the leading columns
  
```

```{r}
# read the unemployment data and tidy it
unemployment_data = read_csv("unemployment.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    -year, names_to = "month", values_to = "unemployment_rate"  
    #except year variable, tidy names into month variable and values into unemployment variable
  ) %>%
  mutate(
    month = str_to_title(month), # capitalize the first letter of the month to match other datasets
    month = month.name[match(month, month.abb)]  # e.g. let "Jan" -> "January"
  )

# merge the datasets
merged_data = pols_data %>%
  left_join(snp_data, by = c("year", "month")) %>%
  left_join(unemployment_data, by = c("year", "month"))

```

```{r}
# export the cleaned datasets and the merged dataset
readr::write_csv(merged_data,   "merged_data.csv")
readr::write_csv(pols_data,     "pols_data.csv")
readr::write_csv(snp_data,      "snp_data.csv")
readr::write_csv(unemployment_data,  "unemployment_data.csv")
```

Description: 
pols_data.csv has 822 rows × 9 columns (year 1947-2015) with monthly counts of officeholders by party: gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem, plus the president variable indicates whether the president was democratic or republican on the associated date (dem = democratic; gop = republican). snp_data.csv has 787 × 3 (year 1950–2015) with year, month, close (the closing values of the S&P stock index). unemployment_data.csv has 816 × 3 (year 1948–2015) with year, month, unemployment (percentage of unemployment on the associate date). After merging all data sets, merged_data.csv contains 822 rows × 11 columns, spanning 1947–2015. Key variables are year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem, president, close, and unemployment_rate (%). 

# Problem 2
```{r}
# read the Mr. Trash Wheel data
mr = read_excel("202509 Trash Wheel Collection Data.xlsx", 
                sheet = "Mr. Trash Wheel", 
                range = "A2:N710", 
                col_names = TRUE,
                na = c(" ", ".", "NA")) %>% 
  janitor::clean_names() %>%  # use reasonable variable names
  filter(!is.na(dumpster)) %>%  # omit rows that do not include dumpster-specific data
  mutate(date = as.Date(date),
         year = as.integer(year),
         month = as.character(month),
         sports_balls = as.integer(round(as.numeric(sports_balls))),  
         #round the number of sports balls to the nearest integer and converts the result to an integer variable
         wheel = "Mr. Trash Wheel"  #keep track of Mr. Trash Wheel
         )  

# read the Professor Trash Wheel data
professor = read_excel("202509 Trash Wheel Collection Data.xlsx", 
                       sheet = "Professor Trash Wheel", 
                       range = "A2:M135", 
                       col_names = TRUE,
                       na = c(" ", ".", "NA")) %>% 
  janitor::clean_names() %>%  # use reasonable variable names
  filter(!is.na(dumpster)) %>%  # omit rows that do not include dumpster-specific data
  mutate(date = as.Date(date),
         year = as.integer(year),
         month = as.character(month),
         wheel = "Professor Trash Wheel")  #keep track of Professor Trash Wheel

# read the Gwynnda Trash Wheel data
gwynnda = read_excel("202509 Trash Wheel Collection Data.xlsx", 
                     sheet = "Gwynns Falls Trash Wheel", 
                     range = "A2:L352", 
                     col_names = TRUE,
                     na = c(" ", ".", "NA")) %>% 
  janitor::clean_names() %>%  # use reasonable variable names
  filter(!is.na(dumpster)) %>%  # omit rows that do not include dumpster-specific data
  mutate(date = as.Date(date),
         year = as.integer(year),
         month = as.character(month),
         wheel = "Gwynnda Trash Wheel")  #keep track of Gwynnda Trash Wheel

# combine the datasets
trash = bind_rows(mr, professor, gwynnda)

# quick stats for inline use
n_observations  = nrow(trash)
n_variables = ncol(trash)
year_min = min(trash$year, na.rm = TRUE)
year_max = max(trash$year, na.rm = TRUE)

prof_total_weight = trash %>%
  filter(wheel == "Professor Trash Wheel") %>%
  summarize(total_tons = sum(weight_tons, na.rm = TRUE)) %>%
  pull()

gwy_june_2022_cigs <- trash %>%
  filter(wheel == "Gwynnda Trash Wheel", year == 2022, str_starts(month, "Jun")) %>%
  summarize(total_cigs = sum(cigarette_butts, na.rm = TRUE)) %>%
  pull()
```

```{r}
# export the cleaned trash dataset
readr::write_csv(trash, "trash.csv")
```
The combined Trash Wheel data set contains `r n_observations` observations and `r n_variables` variables, during `r year_min`–`r year_max`. Each row is a dumpster load with date, month, year, total weight_tons, volume_cubic_yards, and item counts such as plastic_bottles, polystyrene, cigarette_butts, glass_bottles, plastic_bags, wrappers, and sports_balls, plus a wheel identifier indicating Mr., Professor, or Gwynnda. For the available data, Professor Trash Wheel collected `r round(prof_total_weight, 2)` tons of trash in total, and Gwynnda collected `r scales::comma(gwy_june_2022_cigs)` cigarette butts in June 2022.

# Problem 3
```{r}
# read the dataset
zip_info = read.csv("Zip Codes.csv", na = c("NA", ".", ""))%>% 
  janitor::clean_names() # clean names
zori = read.csv("Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", na = c("NA", ".", ""))%>% 
  janitor::clean_names() # clean names

# tidy zori
zori_new = zori %>% pivot_longer(
  x2015_01_31:x2024_08_31, 
  names_to = "date",
  values_to = "ZORI",
  names_prefix = "x"
  ) %>%  # tidy the dates
  mutate(date = ymd(str_replace_all(date, "_", "-")),
         zip_code  = as.integer(region_name)
         ) 

# collapse zip_info to one row per ZIP to avoid many-to-many
zip_info_one = zip_info %>%
  mutate(county = if_else(county == "New York", "Manhattan", county),
         zip_code = as.integer(zip_code)) %>%
  group_by(zip_code) %>%
  summarise(borough = first(county),
            neighborhood = str_c(sort(unique(neighborhood)), collapse = "; "),
            .groups = "drop")

#merge together
NY_zori = zori_new %>%
  left_join(zip_info_one,
            by = "zip_code") %>%
  arrange(zip_code, date) %>%
  relocate(zip_code, date, ZORI)
```

```{r}
# export the merged NY_zori dataset
readr::write_csv(NY_zori, "NY_zori.csv")
```

```{r}
# briefly describe the resulting tidy dataset
n_obs = nrow(NY_zori)
n_zip = dplyr::n_distinct(NY_zori$zip_code)
n_neighborhood = dplyr::n_distinct(NY_zori$neighborhood, na.rm = TRUE)
```
The resulting tidy dataset contains a large number of observations spread across all available months and ZIP codes in New York City. In total, it includes `r n_obs` observations, covering `r n_zip` unique ZIP codes and representing `r n_neighborhood` distinct neighborhoods. This structure ensures that each row corresponds to a unique combination of ZIP code and date, with associated information on rental prices and location.

```{r}
# ZIP codes appear in the ZIP code dataset but not in the Zillow Rental Price dataset
zips_in_zori = NY_zori %>% distinct(zip_code)
zips_only_in_zipfile = zip_info_one %>%
  distinct(zip_code, borough, neighborhood) %>%
  anti_join(zips_in_zori, by = "zip_code") %>%
  arrange(borough, zip_code)

# show a few illustrative examples
head(zips_only_in_zipfile, 5)
```
I found out that there are several ZIP codes that appear in the New York City ZIP code dataset but do not appear in the Zillow rental price dataset. This may be because Zillow only includes areas with sufficient rental listings to generate reliable estimates. For example, 10464 (Southeast Bronx) is a small, low-density residential area with limited rental activity, while 10474 (Hunts Point and Mott Haven) is primarily industrial, with very few housing units. Similarly, 10499 is a non-standard ZIP code used mainly for postal facilities rather than residential addresses. Because these ZIPs lack enough rental data or represent special-purpose areas, Zillow excludes them from its published ZORI dataset.

```{r}
# pull Jan 2020 and Jan 2021 ZORI
zori_2020_01 = NY_zori %>%
  filter(date == "2020-01-31") %>%
  select(zip_code, zori_2020_01 = ZORI)

zori_2021_01 = NY_zori %>%
  filter(date == "2021-01-31") %>%
  select(zip_code, zori_2021_01 = ZORI)

# combine and compute change
zori_20_21 = NY_zori %>%
  distinct(zip_code, borough, neighborhood) %>%
  left_join(zori_2020_01, by = "zip_code") %>%
  left_join(zori_2021_01, by = "zip_code") %>%
  mutate(change_2021_vs_2020 = zori_2021_01 - zori_2020_01) %>%
  arrange(change_2021_vs_2020)

# top 10 largest drops
top10_drop = zori_20_21 %>%
  slice_min(order_by = change_2021_vs_2020, n = 10, with_ties = FALSE) %>%
  select(zip_code, borough, neighborhood, zori_2020_01, zori_2021_01, change_2021_vs_2020)

# pretty table
knitr::kable(top10_drop, digits = 2,
             col.names = c("ZIP", "Borough", "Neighborhood",
                           "ZORI Jan 2020", "ZORI Jan 2021", "Change (2021 - 2020)"))
```
During the COVID-19 pandemic(Jan, 2020 - Jan, 2021), the largest rent declines are overwhelmingly concentrated in Manhattan (New York County), e.g., Lower Manhattan (10007, 10004, 10038), the Lower East Side (10002, 10009), Chelsea/Clinton (10001), Gramercy/Murray Hill (10016, 10010), and Greenwich Village/Soho (10012). These ZIPs show drops of roughly $700–$900 in ZORI. By contrast, many outer-borough areas saw smaller declines (or relative stability), suggesting a shift in demand away from Manhattan’s core during 2020–2021.